---
title: 'Assignment 3: Task 2 - Agglomerative hierarchical clustering'
author: "Sophia Leiker"
date: "2/16/2022"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
library(here)
```

## 1. Introduction

This report will use hierarchical clustering by complete linkage to create a dendrogram showing multivariate clustering for water chemistry by site


**Data Source:** SBC LTER: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000
Creators: Santa Barbara Coastal LTER, & Melack, John M
Citation: Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. [Environmental Data Initiative](https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174). Meta data including metrics used for each variable and data collection methods can be found [here](https://portal.edirepository.org/nis/metadataviewer?packageid=knb-lter-sbc.6.16). This data package contains stream water chemistry measurements taken in Santa Barbara area watersheds, beginning in 2000. Stream water samples are collected weekly during non-storm flows in winter, and bi-weekly during summer.

Some variables of interest include:

- Total Dissolved Nitrogen uM
- Total Dissolved Phosphorus uM
- Total Particulate Carbon micromolesPerLiter
- Total Particulate Nitrogen micromolesPerLiter
- Total Suspended Solids (milligramPerLiter)


```{r}
# Reading in the data
stream <- read_csv(here("data", "sbc_lter_registered_stream_chemistry.csv"))
```


***
## 2. Data Wrangling

- Convreting -999 values to NA
- 
```{r}
stream[stream==-999]<-NA

stream_site <- stream %>% 
  group_by(site_code) %>% 
  summarize()
  
```


***

## 3. Dendrogram





```{r}
# Reading in the data
world <- read_csv(here("data", "world_env_vars.csv"))
```


***

## 4. Data Citation

Data source: Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. [Environmental Data Initiative](https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174)

## END TASK




