---
title: 'Assignment 3: Task 1 - Principal Components Analysis'
author: "Sophia Leiker"
date: "2/15/2022"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(janitor)

```

## 1. Introduction

This report will carry out a Principal Components Analysis and will create a biplot for result interpretation. 
The analysis will be performed on environmental and climatic variables at the country level. 


**Data Source:** The data used  was compiled and provided by @zander_venter on Kaggle. “This data is all acquired through [Google Earth Engine](https://earthengine.google.com/) where publicly available remote sensing datasets have been uploaded...Most of the data is derived by calculating the mean for each country at a reduction scale of about 10km.” 


```{r}
# Reading in the data
world <- read_csv(here("data", "world_env_vars.csv"))

```

***

